Improving the portuguese hyphenation rules

### 2023-11-24 article/article.tex
fix Issue #35
fix Issue #37
fix Issue #40

In Portuguese, hyphenations are allowed on syllables boundaries and, in
general, follow phonological principles.  According to the Grammar
\parencite{cunha2016,bergstrom2011,cegalla2008}, some rules might still apply:
\begin{enumerate}
    \item\label{rule-di-triphthong} diphthong or triphthong should not be split
	(e.g. \emph{m\underline{ui}-to}, \emph{Pa-ra-g\underline{uai}});
    \item\label{rule-unstressed} the sequences \emph{ia}, \emph{ie}, \emph{io}, \emph{oa}, \emph{ua},
	\emph{ue} and \emph{uo}, when in final unstressed position, should not be split 
	(e.g. \emph{gló-r\underline{ia}}, \emph{vi-tó-r\underline{ia}}, 
	\emph{cá-r\underline{ie}}, \emph{es-pé-c\underline{ie}}, 
	\emph{Má-r\underline{io}}, \emph{má-g\underline{oa}}, 
	\emph{ré-g\underline{ua}}, \emph{tê-n\underline{ue}}, 
	\emph{con-tí-g\underline{uo}}, \emph{am-bí-g\underline{uo}});
    \item\label{rule-c-clusters} consonant clusters starting a syllable should not be
	split (e.g. \emph{\underline{pn}eu-má-ti-co}, \emph{\underline{ps}i-có-lo-go}, 
	\emph{\underline{mn}e-mô-ni-co});
    \item\label{rules-digraphs-ns} the digraphs	\emph{ch}, \emph{lh}, \emph{nh}
	should not be split (e.g. \emph{ra-\underline{ch}ar},
	\emph{a-bro-\underline{lh}os}, \emph{ma-\underline{nh}ã};
    \item\label{rule-guqu} bigrams like \emph{gu} and \emph{qu} whose vowel \emph{u} is not 
	pronounced are never separated from the vowel or diphthong that follows it 
	(e.g. \emph{ne-\underline{gue}}, \emph{pe-\underline{que}});
    \item\label{rule-nasalization} since they are digraphs, a vowel and its following 
	nasalization marker (a graphic nasal consonant) should not be split (e.g. \emph{\underline{am}-bição},
	\emph{m\underline{an}-cha});
    \item\label{rule-decreasing} decreasing diphthongs should not be split (e.g. 
	\emph{\underline{ai}-ro-so}, \emph{ca-d\underline{ei}-ra}, 
	\emph{o-ra-ç\underline{ão}});
    \item\label{rule-rising} rising diphthong should not be split (e.g.
	\emph{a-b\underline{ai}-xo}; \emph{c\underline{au}-te-la}, 
	\emph{pa-p\underline{éi}s}, \emph{cha-p\underline{éu}}, 
	\emph{pre-f\underline{ei}-to}, \emph{r\underline{eu}-nir},
	\emph{n\underline{oi}-te}; \emph{ca-la-b\underline{ou}-ço}, 
	\emph{as-te-r\underline{ói}-de}; 
	\emph{re-tri-b\underline{ui}});
    \item\label{rule-singlev} disyllables whose syllable has a single vowel should 
	not be split (e.g. \emph{\underline{a}to}, \emph{ru\underline{a}}, 
	\emph{\underline{ó}dio}, \emph{\underline{u}nha});
    \item\label{rule-orphan} words with more than two syllables, when divided, cannot 
	isolate a syllable composed of a single vowel (e.g. \emph{\underline{a}gos-to}, 
	\emph{la-go\underline{a}}, \emph{\underline{i}da-de});

@@

Rules \ref{rule-singlev} and \ref{rule-orphan} are primarily aimed at ensuring
proper readability of the text, aligning with \TeX{} approach to deal with
widows and orphans. As mentioned in \Cref{sec-intro}, the variables
\verb|lefthyphenmin| and \verb|righthyphenmin| are used to control the minimum
length for fragments of hyphenated words. For that reason, rules
\ref{rule-singlev} and \ref{rule-orphan} are fiddling rules in \TeX{}
hyphenation, and that is not reason to take than in account in our analysis to
improve Portuguese hyphenation rules. Furthermore, get the full hyphenation of
words might be useful for text-to-speech applications \cite{libossek2000,trogkanis2010}.

@@

In some dictionaries, words are usually marked with a colon (`:') to indicate a
dubious ending, classifying them as `apparent proparoxytones'. In Portuguese,
words ending in \emph{ea}, \emph{eo}, \emph{ia}, \emph{ie}, \emph{io},
\emph{ua}, \emph{ue}, or \emph{uo}, with the stressed syllable before these
endings, can be considered either paroxytones ending in a rising diphthong or
proparoxytones ending in a hiatus. The words \emph{his-tó-ri:a} and \emph{on-du-la-tó-ri:o}
are examples of words marked as apparent proparoxytones, meaning they might
be regarded as paroxytones \emph{his-tó-ria} and \emph{on-du-la-tó-rio} or
as proparoxytones \emph{his-tó-ri-a} and \emph{on-du-la-tó-ri-o}.
The script \texttt{hyphenationagreements.sh} might deal with these apparent proparoxytones
in four different ways: we might keep the hyphenations with the colon mark to
indicate an apparent proparoxytones; or we might only the paroxytones or only the proparoxytones
counterpart; or we may count both paroxytones and proparoxytones.
As there is no consensus in the literature, we have decided to count both
hyphenations. In our hyphenation dictionary, we have found \NumberOfAppProp{} words
marked as apparent proparoxytones.

>>>>

In Portuguese, hyphenations are allowed on syllables boundaries and, in
general, follow phonological principles.  According to the Grammar
\parencite{cunha2016,bergstrom2011,cegalla2008}, some rules might still apply:
\begin{enumerate}
    \item\label{rule-di-triphthong} diphthong or triphthong should not be split
	(e.g. \emph{m\underline{ui}-to}, \emph{Pa-ra-g\underline{uai}});
    \item\label{rule-unstressed} the sequences \emph{ia}, \emph{ie}, \emph{io}, \emph{oa}, \emph{ua},
	\emph{ue} and \emph{uo}, when in final unstressed position, should not be split 
	(e.g. \emph{gló-r\underline{ia}}, \emph{vi-tó-r\underline{ia}}, 
	\emph{cá-r\underline{ie}}, \emph{es-pé-c\underline{ie}}, 
	\emph{Má-r\underline{io}}, \emph{má-g\underline{oa}}, 
	\emph{ré-g\underline{ua}}, \emph{tê-n\underline{ue}}, 
	\emph{con-tí-g\underline{uo}}, \emph{am-bí-g\underline{uo}});
    \item\label{rule-c-clusters} consonant clusters starting a syllable should not be
	split (e.g. \emph{\underline{pn}eu-má-ti-co}, \emph{\underline{ps}i-có-lo-go}, 
	\emph{\underline{mn}e-mô-ni-co});
    \item\label{rules-digraphs-ns} the digraphs	\emph{ch}, \emph{lh}, \emph{nh}
	should not be split (e.g. \emph{ra-\underline{ch}ar},
	\emph{a-bro-\underline{lh}os}, \emph{ma-\underline{nh}ã};
    \item\label{rule-guqu} bigrams like \emph{gu} and \emph{qu} whose vowel \emph{u} is not 
	pronounced are never separated from the vowel or diphthong that follows it 
	(e.g. \emph{U-ru-\underline{guai}}, \emph{pe-\underline{que}});
    \item\label{rule-nasalization} since they are digraphs, a vowel and its following 
	nasalization marker (a graphic nasal consonant) should not be split (e.g. \emph{\underline{am}-bição},
	\emph{m\underline{an}-cha});
    \item\label{rule-decreasing} decreasing diphthongs should not be split (e.g. 
	\emph{\underline{ai}-ro-so}, \emph{ca-d\underline{ei}-ra}, 
	\emph{o-ra-ç\underline{ão}});
    \item\label{rule-rising} rising diphthong should not be split (e.g.
	\emph{a-b\underline{ai}-xo}; \emph{c\underline{au}-te-la}, 
	\emph{pa-p\underline{éi}s}, \emph{cha-p\underline{éu}}, 
	\emph{pre-f\underline{ei}-to}, \emph{r\underline{eu}-nir},
	\emph{n\underline{oi}-te}; \emph{ca-la-b\underline{ou}-ço}, 
	\emph{as-te-r\underline{ói}-de}; 
	\emph{re-tri-b\underline{ui}});
    \item\label{rule-singlev} disyllables whose syllable has a single vowel should 
	not be split (e.g. \emph{\underline{a}to}, \emph{ru\underline{a}}, 
	\emph{\underline{ó}dio}, \emph{\underline{u}nha});
    \item\label{rule-orphan} words with more than two syllables, when divided, cannot 
	isolate a syllable composed of a single vowel (e.g. \emph{\underline{a}gos-to}, 
	\emph{la-go\underline{a}}, \emph{\underline{i}da-de});

@@

Rules \ref{rule-singlev} and \ref{rule-orphan} are primarily aimed at ensuring
proper readability of the text, aligning with \TeX{} approach to deal with
widows and orphans. As mentioned in \Cref{sec-intro}, the variables
\verb|lefthyphenmin| and \verb|righthyphenmin| are used to control the minimum
length for fragments of hyphenated words. For that reason, rules
\ref{rule-singlev} and \ref{rule-orphan} are fiddling rules in \TeX{}
hyphenation, and that is not reason to take than in account in our analysis to
improve Portuguese hyphenation rules. Furthermore, get the full hyphenation of
words might be useful for text-to-speech applications \parencite{libossek2000,trogkanis2010}.

@@

In certain dictionaries, words are commonly annotated with a colon (`:') to
signify a potential variation in pronunciation or production, specifically as a
diphthong or hiatus, classifying them as `apparent proparoxytones'.  In
Portuguese, words ending in \emph{ea}, \emph{eo}, \emph{ia}, \emph{ie},
\emph{io}, \emph{ua}, \emph{ue}, or \emph{uo}, with the stressed syllable
before these endings, can be considered either paroxytones ending in a rising
diphthong or proparoxytones ending in a hiatus. The words \emph{his-tó-ri:a}
and \emph{on-du-la-tó-ri:o} are examples of words marked as apparent
proparoxytones, meaning they might be regarded as paroxytones \emph{his-tó-ria}
and \emph{on-du-la-tó-rio} or as proparoxytones \emph{his-tó-ri-a} and
\emph{on-du-la-tó-ri-o}.  The script \texttt{hyphenationagreements.sh} might
deal with these apparent proparoxytones in four different ways: we might keep
the hyphenations with the colon mark to indicate an apparent proparoxytones; or
we might only the paroxytones or only the proparoxytones counterpart; or we may
count both paroxytones and proparoxytones.  As there is no consensus in the
literature, we have decided to count both hyphenations. In our hyphenation
dictionary, we have found \NumberOfAppProp{} words marked as apparent
proparoxytones.
